Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-11-30 threshold/draft.Rnw
sequential filter option

%%
%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'biclust' package installed in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.
%%
\documentclass[10pt]{article}

@@

We illustrate that step 4 of Algorithm \ref{prune} can be computed in parallel
by comparing the above results run on our reference test PC with results run on
a modest GNU/Linux cluster at Paradigm4, Inc. The cluster consisted of four
computers connected by 10 gigabit ethernet. Each computer contained two Intel
Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per computer) and 64$\,$GB
ECC registered RAM running at 1,600$\,$MHz running CentOS 6.6 and 64-bit R
version 3.2.2 with the reference R BLAS library.

No code change was required to run in parallel on the cluster. The reference R
{\tt tcor()} implementation uses R's {\tt foreach}\cite{foreach} framework to
run on arbitrary parallel configurations (``back ends''). We simply registered
a {\tt doRedis}\cite{doredis} parallel back end with 16 R workers per computer
prior to invoking {\tt tcor()}.

The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
below.

<<echo=FALSE, results=tex>>=
# Default values:
t1=10449.670
t2=NA
t3=NA
m1=1261.6
m2=NA
m3=NA

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  # simulated data, replicate rows 27x
}
require(xtable)
ans = data.frame(Method=c("tcor (4 core PC)","brute force","tcor (64 core cluster)"), `Wall-clock time (s)`=c(t1, t2, t3), `Peak excess memory use (MB)`=c(m1, m2, m3), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99)", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="NA")
@

@@

%% IRLBA implementations and supplemental code
\bibitem{irlbar}
Lewis, Bryan,  \url{https://cran.r-project.org/web/packages/irlba}.

\bibitem{sup}
Lewis, Bryan,  \url{https://github.com/bwlewis/tcor}.

@@

>>>>

%
%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'biclust' package installed in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.
%%
\documentclass[10pt]{article}

@@

We illustrate that steps 4--7 of Algorithm \ref{prune} can be computed in
parallel by comparing the above results run on our reference test PC with
results run on a modest GNU/Linux cluster at Paradigm4, Inc. The cluster
consisted of four computers connected by 10 gigabit ethernet. Each computer
contained two Intel Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per
computer) and 64$\,$GB ECC registered RAM running at 1,600$\,$MHz running
CentOS 6.6 and 64-bit R version 3.2.2 with the reference R BLAS library.

No code change was required to run in parallel on the cluster. The reference R
{\tt tcor()} implementation uses R's {\tt foreach}\cite{foreach} framework to
run on arbitrary parallel configurations (``back ends''). We simply registered
a {\tt doRedis}\cite{doredis} parallel back end with 16 R workers per computer
(64 total R worker processes) prior to invoking {\tt tcor()}.

The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
below. The cluster example computed the filtering step 7 of Algorithm
\ref{prune} in parallel across the 64 remote R processes. That approach incurs,
in this example, an additional memory overhead of about 16$\,$GB because each
remote R process requires access to a copy of the data matrix $A$. More thrifty
adaptations of the algorithm are possible and could, for example, share the
memory required by $A$ between worker R processes on each computer.

<<echo=FALSE, results=tex>>=
# Default values:
t1=10449.670
t3=482
t4=NA
m1=1261.6
m3=994.3
m4=42282.4

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
 # methylation data
}
require(xtable)
ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall-clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="")
@
From a set of almost 400 thousand vectors, Algorithm \ref{prune} finds 913,601
pairs of vectors with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm also runs well in constrained
memory settings.

@@

%% doredis
\bibitem{doredis}
Lewis, Bryan. The doRedis package for R. \url{https://cran.r-project.org/web/packages/doRedis}.

%% IRLBA implementations and supplemental code
\bibitem{irlbar}
Lewis, Bryan. The irlba package for R.  \url{https://cran.r-project.org/web/packages/irlba}.

\bibitem{sup}
Lewis, Bryan. The tcor package for R.  \url{https://github.com/bwlewis/tcor}.

@@

%% foreach
\bibitem{foreach}
Weston, Steve. The foreach package for R. \url{https://cran.r-project.org/web/packages/foreach/vignettes/foreach.pdf}.
